Did WW2 Produce an Equivalent to a Modern Computer?

### 2017-06-04 Draft 2/Section 4 - Zuse's Z3.tex
dEaDlInE ;iN 1~5 .dAyS wHaT'

At this point, we have an almost complete definition of 'computer' - a machine which is able to fully implement the concept of effective computability. This can be done most simply through the implementation of an effectively computable method of computation, such as Turing Machines. However, in order to address the question of whether this was achieved during WW1 we must apply this test to the physical instruments created within this time.

\textit{TODO: Colossus}

@@

In this section I shall give a broad functional overview of the Z3. I shall explore primarily the function and usage of the machine, the aspects that would be visible to the programmer. I shall not go into detail about the operation of the machine itself unless it directly influences the machine's usage. The description of the machine was derived from \cite{rojas1997z3architecture}.

The Z3 was primarily purposed towards performing simple arithmetical calculations - addition, subtraction, multiplication, division, and the square root operation. The operands would be read from memory or from input from a keyboard, and the results would be written to memory or shown on a display. Programs would be written as sequences of 8-bit instructions on punched tape, and would be performed as they are read by the machine.

@@

The two memory instructions, Pr \textit{z} and Ps \textit{z}, load and store respectively the value of registers from and to memory. Pr \textit{z} will transfer the data stored at address \textit{z} to $R_1$ if it is clear, otherwise it will be written to $R_2$, overwriting any previous data. Ps \textit{z} transfers the value of $R_1$ only to address \textit{z}, and clears both $R_1$ and $R_2$ (such that the next Pr \textit{z} operation will write to $R_1$).

@@

For the sake of simplicity, in this section we shall use a more intuitive form to express instructions for the machine. We shall write $a = b\ \mathrm{op}\ c$ to simplify statements of the form:s
\begin{gather*}
\begin{aligned}
&\mathrm{Pr}\ b\\
&\mathrm{Pr}\ c\\
&\textrm{op}\\ 
&\mathrm{Ps}\ a\\
\end{aligned}
\end{gather*}

@@

Firstly and fundamentally we shall construct conditional branches using a technique shown in \cite{ibarra1983control} which allows the implementation of a form of case statement as found in modern programming languages such as JavaScript. Such an operation performs one of a set of different operations depending on the value of a given variable.

Given a positive integer variable $z$, the variable on which the outcome of the statement depends and the set of operations to be performed depending on the value of $z$, $P_1, ... , P_n$, we can produce a sequence of operations that will allow only the operation $P_z$ to perform. Due to the nature of the Z3 requiring a full sequence of commands to be executed without exception, every one of these operations will be evaluated; however, only operation $P_z$ will be allowed to actually affect memory.

@@

Or, for clarity, $a = (a \cdot t) + (1 - t)(b\ \mathrm{op}\ c)$. If $t = 1$, $v$ is set to $t$; otherwise it is set to $0$. Furthermore, if $t = 0$, $u$ is set to the outcome of the operation; otherwise it is set to $0$ again. Therefore, if $t = 0$, $a$ is set to the outcome of the operation; otherwise it is set back to its original value. As such, after this transformation has been put in place for every $P$ operation, and a method to set $t$ at the start of the case statement has been implemented, every new command in this program can be executed - most superfluously - in order to only use the $z$-th command. Next we must produce a set of operations that will perform the computation
\begin{equation*}
\mathrm{if\ (}z = i\mathrm{)\ then\ } t = 0 \mathrm{\ else\ } t = 1
\end{equation*}

@@

Using this method to produce case statements, we can produce an equivalent to an indirect addressing operation. This operation is of form $n = (x)$, which sets $n$ to the value at address location $x$, where $x$ is an integer $n_a \leq x \leq n_b$, $n_a$ and $n_b$ being integer constants representing the limits of the region of memory that can be addressed by this function. This function is equivalent to:

\subsubsection{State Storage \& Access}

With the Z3's two new-found tools, production of a universal machine is incredibly simple. The machine would first be prepared with the initial state (an integer) at the address $s$, and the initial index of the cursor (also an integer, $a$ by default) at the address $h$. Then the following program would be run repeatedly with the ends of the tape attached to each other.
\begin{gather*}
\begin{aligned}
&r = (h)						\\
&p = k 							\\
&\mathrm{case}\ (s)\ \{			\\
&\qquad	... ,					\\
&\qquad	n:\ w = [symbol],		\\
&\qquad	...						\\
&\}								\\
&\mathrm{case}\ (s)\ \{			\\
&\qquad	...,					\\
&\qquad	n:\ t = [-1/0/1],		\\
&\qquad	...						\\
&\}								\\
&h = h + t						\\
&\mathrm{case}\ (s)\ \{			\\
&\qquad ...,					\\
&\qquad	n:\ s = [new\ state],	\\
&\qquad	...						\\
&\}								\\
\end{aligned}

\end{document}

>>>>

At this point, we have an almost complete definition of 'computer' - a machine which is able to fully implement the concept of effective computability. This can be done most simply through the implementation of an effectively computable method of computation, such as a Turing Machine. In order to address the question of whether this was achieved during WW1 we must apply this test to the physical instruments created during this period.
\textit{TODO: Colossus}

@@

In this section I shall give a broad functional overview of the Z3. I shall explore primarily the function and usage of the machine, i.e. the aspects that would be visible to the programmer. I shall not go into detail about the operation of the machine itself unless it directly influences the machine's usage. The description of the machine was derived from \cite{rojas1997z3architecture}.

The Z3 was primarily purposed towards performing simple arithmetical calculations - addition, subtraction, multiplication, division, and square root. The operands would be read from memory or from input from a keyboard, and the results would be written to memory or shown on a display. Programs could be written as sequences of 8-bit instructions on punched tape, and would be performed as they are read by the machine.

@@

The two memory instructions Pr \textit{z} and Ps \textit{z} load and store respectively the value of registers from and to memory. Pr \textit{z} transfers the data stored at address \textit{z} to $R_1$ if $R_1$ is clear; otherwise it is written to $R_2$, overwriting any previous data. Ps \textit{z} transfers the value of $R_1$ only to address \textit{z}, and clears both $R_1$ and $R_2$ (such that the next Pr \textit{z} operation will write to $R_1$).

@@

For the sake of simplicity, in this section we shall use a more intuitive form to express instructions for the machine. We shall write $a = b\ \mathrm{op}\ c$ to simplify statements of the form:
\begin{gather*}
\begin{aligned}
&\mathrm{Pr}\ b\\
&\mathrm{Pr}\ c\\
&\textrm{op}\\ 
&\mathrm{Ps}\ a\\
\end{aligned}
\end{gather*}

@@

Firstly and fundamentally we shall construct conditional branches using a technique shown in \cite{ibarra1983control} which allows the implementation of a form of switch statement as found in many modern C-like programming languages. Such an operation performs one of a set of different operations depending on the value of a given variable.

Given a positive integer variable $z$ on which the outcome of the statement depends and the set of operations to be performed depending on the value of $z$, $P_1, ... , P_n$, we can produce a sequence of commands that will allow only the operation $P_z$ to perform. Due to the nature of the Z3 requiring a full sequence of commands to be executed without exception, every one of these operations will be evaluated; however, only operation $P_z$ will be allowed to actually affect memory.

@@

Or, for clarity, $a = (a \cdot t) + (1 - t)(b\ \mathrm{op}\ c)$. If $t = 1$, $v$ is set to $t$; otherwise it is set to $0$. Furthermore, if $t = 0$, $u$ is set to the outcome of the operation; otherwise it is set to $0$ again. Therefore, if $t = 0$, $a$ is set to the outcome of the operation; otherwise it is set back to its original value. As such, after this transformation has been put in place for every $P$ operation, and a method to set $t$ at the start of the switch statement has been implemented, every new command in this program can be executed - most superfluously - in order to only use the $z$-th command. Next we must produce a set of operations that will perform the computation
\begin{equation*}
\mathrm{if\ (}z = i\mathrm{)\ then\ } t = 0 \mathrm{\ else\ } t = 1
\end{equation*}

@@

Using this method to emulate switch statements, we can produce an equivalent to an indirect addressing operation. This operation is of form $y = (x)$, which sets $n$ to the value at address location $x$, where $x$ is an integer $n_a \leq x \leq n_b$, $n_a$ and $n_b$ being integer constants representing the limits of the region of memory that can be addressed by this function. $x$ could potentially be the output from a previous arithmetical operation. Relying on the fact that this region is finite (and furthermore that the tape length is also finite), this can be done using a switch statement with a case for every address $z$ between $n_a$ and $n_b$ as shown.
\begin{gather*}
\begin{aligned}
&\bm{\mathrm{switch}}\ (x)\ \{	\\
&\qquad ... ,					\\
&\qquad z:\	\mathrm{Pr}\ z;\ \mathrm{Ps}\ a			\\
&\qquad ... 					\\
&\}								\\
\end{aligned}
\end{gather*}
At this point, it becomes very clear the extent to which the Z3 is not suited to universal computation. When every single applicable memory address must be specified individually, with five operations per address to implement the case, each operation consisting of four instructions, it is obvious that the Z3 is in no way a \textit{practical} universal machine.

To emulate a Turing machine using these new tools, one considers the tape to be the region of memory between $n_a$ and $n_b$, and 
\end{document}
